\documentclass[11pt]{article}

% Change "review" to "final" for the camera-ready version.
% Change to "preprint" for a non-anonymous version with page numbers.
\usepackage[review]{acl}

\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xspace}

\newcommand{\method}{Negation-Contrastive Reranking\xspace}

\title{\method for Information Retrieval with Negated Queries}

\author{Anonymous Authors}

\begin{document}
\maketitle

\begin{abstract}
Negation is a compact linguistic device with large consequences for information retrieval.
Queries such as ``cake recipes without butter'' specify not only a topical intent, but also an exclusion constraint that can be easily missed by lexical and dense retrievers.
We propose a simple reranking framework for negation-sensitive retrieval: retrieve once with the original negated query, retrieve again with an LLM-generated contrastive query in which the negated constraint is replaced by its affirmative counterpart, and rerank documents by the difference between the two retrieval scores.
This paper outlines the task, method, and evaluation protocol for studying whether contrastive score subtraction improves retrieval under negated user intents.
\end{abstract}

\section{Introduction}

Users frequently express information needs through negation, exclusion, and absence.
For example, a user searching for ``cake recipes without butter'' is not merely interested in cake recipes, but in recipes that avoid a specific ingredient.
Standard retrieval models may assign high scores to documents that strongly match the topical terms while failing to enforce the negated constraint.

Recent benchmarks make this failure mode increasingly visible.
NevIR shows that many neural retrievers struggle to distinguish minimally different documents whose relevance flips under negation \citep{weller-etal-2024-nevir}, while ExcluIR broadens the problem to realistic exclusionary queries where users specify what they do not want \citep{zhang-etal-2024-excluir}.
NegConstraint further frames negative constraints as a logical consistency problem and addresses it with a neural-symbolic reranker \citep{xu-etal-2025-negconstraint}.
Together, these studies demonstrate that negation-sensitive retrieval is not solved by stronger semantic matching alone.
However, existing solutions largely emphasize new benchmarks, additional negation-specific training data, or specialized model architectures.
This leaves open a practical question: can an existing retriever be made more sensitive to negated intent at inference time, without retraining or changing its scoring function?

We study a lightweight approach to negation-sensitive reranking.
Given a negated query, we use a large language model (LLM) to generate a contrastive affirmative query.
For example, ``cake recipes without butter'' is transformed into ``cake recipes with butter''.
We then compare the score of each document under the original query and the contrastive query.
Documents that are relevant to the original query but also strongly match the contrastive query are penalized.
The key intuition is that the affirmative rewrite approximates the forbidden intent.
Thus, subtracting its retrieval score turns the retriever's own preference for the opposite condition into a penalty signal.
Unlike fine-tuning-based approaches, this mechanism is model-agnostic and can be layered over sparse, dense, or neural reranking systems as long as they expose comparable scores.

Our intended contributions are:
\begin{itemize}
    \item We formulate negated-query retrieval as a contrastive reranking problem.
    \item We propose a simple score-difference reranker that can be applied on top of existing sparse or dense retrievers.
    \item We outline an evaluation protocol for measuring retrieval quality on queries containing explicit negation.
\end{itemize}

\section{Task Definition}

Let $q^{-}$ denote an original user query containing a negated constraint, such as ``cake recipes without butter''.
Let $q^{+}$ denote a contrastive query generated by replacing the negated component with an affirmative counterpart, such as ``cake recipes with butter''.
Given a corpus $\mathcal{D}$, the goal is to rank documents $d \in \mathcal{D}$ such that documents satisfying the intent of $q^{-}$ are ranked above documents that satisfy only the broad topic or the opposite constraint.

\section{Method}

\subsection{Contrastive Query Generation}

We prompt an LLM to identify the negated span in $q^{-}$ and rewrite it as an affirmative counterpart while preserving the rest of the query intent.
The rewrite should change only the polarity-sensitive component.
Table~\ref{tab:rewrite-examples} shows examples.

\begin{table}[t]
\centering
\small
\begin{tabular}{ll}
\toprule
Original query $q^{-}$ & Contrastive query $q^{+}$ \\
\midrule
cake recipes without butter & cake recipes with butter \\
hotels with no parking fee & hotels with parking fee \\
non-alcoholic cocktails & alcoholic cocktails \\
\bottomrule
\end{tabular}
\caption{Examples of LLM-generated contrastive queries.}
\label{tab:rewrite-examples}
\end{table}

\subsection{Score-Difference Reranking}

We retrieve or score documents with both the original query and the contrastive query.
Let $s(d, q)$ be the retrieval score assigned to document $d$ for query $q$.
The final reranking score is:

\begin{equation}
    s_{\mathrm{final}}(d, q^{-}) =
    s(d, q^{-}) - \lambda s(d, q^{+}),
    \label{eq:score-difference}
\end{equation}

where $\lambda$ controls the strength of the contrastive penalty.
When $\lambda=1$, the model directly subtracts the contrastive score from the original score.
Documents that match the affirmative counterpart receive a lower final score.

\subsection{Candidate Pooling}

We consider two candidate construction strategies:
\begin{enumerate}
    \item Retrieve top-$k$ documents with $q^{-}$ and rerank only those documents.
    \item Retrieve top-$k$ documents with both $q^{-}$ and $q^{+}$, merge the candidate sets, and rerank the union.
\end{enumerate}

\section{Experimental Setup}

\subsection{Datasets}

We will evaluate on retrieval queries containing explicit negation.
Candidate sources include existing IR benchmarks with manually identified negated queries, synthetically generated negation queries, or domain-specific datasets such as recipes, products, and travel search.

\subsection{Baselines}

We compare against:
\begin{itemize}
    \item retrieval with the original query only;
    \item retrieval with query expansion using the LLM rewrite;
    \item cross-encoder reranking, when available;
    \item ablations without score subtraction or without LLM rewriting.
\end{itemize}

\subsection{Metrics}

We report standard retrieval metrics such as nDCG@$k$, MRR@$k$, Recall@$k$, and Precision@$k$.
We also report constraint-satisfaction accuracy, measuring whether top-ranked documents obey the negated condition.

\section{Analysis Plan}

We analyze when score subtraction helps or hurts by grouping queries according to negation type, domain, rewrite quality, and retriever family.
We also inspect failure cases where the LLM rewrite changes the topic, where the retriever assigns poorly calibrated scores across queries, or where the corpus does not explicitly state the negated property.

\section{Related Work}

Recent work shows that negation remains a brittle phenomenon for neural information retrieval.
\citet{weller-etal-2024-nevir} introduce NevIR, a targeted benchmark in which retrieval models must distinguish between document pairs that differ primarily in negation.
Their results show a clear architectural gap: cross-encoders handle negation better than late-interaction, bi-encoder, and sparse neural models, but even strong models remain far from human performance.
This motivates treating negation not as a minor lexical variant, but as a semantic constraint that can reverse relevance judgments.

Exclusionary retrieval extends this concern from controlled contrastive pairs to realistic user queries.
\citet{zhang-etal-2024-excluir} propose ExcluIR, a benchmark and training resource for queries that specify what users do not want.
They find that existing retrievers struggle with exclusionary intent, while training on exclusionary examples improves performance without closing the gap to human judgments.
Similarly, \citet{xu-etal-2025-negconstraint} study negative-constraint queries through NegConstraint and propose NS-IR, a neural-symbolic method that uses first-order logic, logic alignment, and connective constraints to improve consistency between queries and documents.
These datasets and methods establish that explicit negative constraints are a distinct retrieval challenge, not merely a special case of ordinary topical matching.

Our work is closest in spirit to this line of negation-sensitive and exclusionary IR, but differs in its intervention point.
Rather than constructing a new retriever or fine-tuning on negation-specific training data, we use an LLM to generate an affirmative counterpart to the original negated query and apply score-difference reranking at inference time.
This connects negation-aware retrieval with query rewriting and contrastive reranking: the contrastive query functions as an estimate of the opposite intent, allowing documents that match the forbidden condition to be penalized while preserving compatibility with existing sparse, dense, or neural rerankers.

\section{Conclusion}

We proposed a contrastive reranking framework for information retrieval with negated queries.
By comparing retrieval scores from an original negated query and an LLM-generated affirmative counterpart, the method aims to penalize documents that satisfy the opposite constraint.
Future work will evaluate the approach across retrievers, domains, and negation types.

\section*{Limitations}

The method depends on the quality of LLM-generated rewrites and assumes that scores across the original and contrastive queries are comparable.
It may be less effective when documents imply constraints indirectly or when the negated attribute is absent from the document text.

\section*{Ethics Statement}

The proposed method is intended to improve faithful retrieval for user-specified constraints.
Potential risks include over-penalizing relevant documents due to erroneous rewrites or retrieval score miscalibration.

\bibliography{custom}

\appendix

\section{Prompt Template}

\begin{quote}
\small
Given a user query containing negation, rewrite only the negated condition into its affirmative counterpart while preserving the rest of the query. Return only the rewritten query.
\end{quote}

\end{document}
